Early warning signals: The charted and uncharted territories

### 2013-05-20 manuscript.tex
Compiled vanilla tex version and pdf, looks great! closes #26

\begin{figure}[htbp]
\centering
\includegraphics{ews-venn.eps}
\caption{Venn diagram representing the intersecting domains of rapid
regime shifts, bifurcations, and critical slowing down. Labels in italic
are example phenomena that occur in each domain. Roman numerals indicate
example literature (right) exploring each domain, and also refer to
sections below describing those domains. The center domain (I) where all
three phenomena intersect, is the most extensively researched domain of
the EWS field. Literature outside this charted region does not yet
provide the needed EWS, but hints where existing signals based on CSD
may be insufficient or misleading.}
\end{figure}

@@

Saddle nodes are only one of a variety of bifurcations, which can cause
rapid changes in system dynamics. Other bifurcations can cause long-term
changes in system dynamics without a gradual pass through a state with
zero eigenvalue, and therefore, not exhibit CSD. Many of these examples
can in fact show patterns in typical early warning indicator variables
such as variance or autocorrelation that are completely opposite to the
patterns seen in the saddle-node case.

@@

Some rapid regime shifts are not due to bifurcations at all. A large
external forcing (as illustrated in Figure 4) may change the behavior of
a system without any warning. This mechanism is commonly recognized,
(Scheffer et al. 2001, 2009, Barnosky et al. 2012, Scheffer et al.
2012), but others are possible. An internal stochastic event may switch
a system between dynamic regimes, or a change in system behavior may be
the manifestation of a long-term transient. In none of these cases would
CSD be expected to precede such changes.

@@

The future of early warning signals lies in the uncharted territory. For
certain classes of transitions, such as stochastically-driven regime
shifts, prediction may not be possible. In such cases, managing for
resilience may be the only option. Likewise, regime shifts driven by
external perturbation or strong forcing are not predictable \emph{if}
the scope of management does not include the external causes. Proper
scoping of the management problem can avoid this situation (Fischer et
al. 2009, Alliance 2010, Polasky et al. 2011).

>>>>

\begin{figure}[htbp]
\centering
\includegraphics{ews-venn.eps}
\caption{Venn diagram representing the intersecting domains of rapid
regime shifts, bifurcations, and critical slowing down. Labels in italic
are example phenomena that occur in each domain. Roman numerals and
indicate example literature (right) exploring each domain, and also
refer to sections below describing those domains. Each dot represents a
study in the domain. Studies and dots in grey represent literature not
explicitly testing EWS, but which demonstrate phenomena related to EWS.
The center domain (I) where all three phenomena intersect, is the most
extensively researched domain of the EWS field. Literature outside this
charted region does not yet provide the needed EWS, but hints where
existing signals based on CSD may be insufficient or misleading.}
\end{figure}

@@

Saddle nodes are only one of a variety of bifurcations, which can cause
rapid changes in system dynamics. Other bifurcations can cause long-term
changes in system dynamics without a gradual pass through a state with
zero eigenvalue, and therefore, not exhibit CSD. Many of these examples
can in fact show patterns in typical early warning indicator variables
such as variance or autocorrelation that are completely opposite to the
patterns seen in the saddle-node case. Several of these examples are
found outside the EWS literature, indicating a need to expand the range
of systems studied for EWS.

@@

Some rapid regime shifts are not due to bifurcations at all. A large
external forcing (as illustrated in Figure 4) may change the behavior of
a system without any warning. This mechanism is commonly recognized,
(Scheffer et al. 2001, 2009, Barnosky et al. 2012, Scheffer et al.
2012), but others are possible. An internal stochastic event may switch
a system between dynamic regimes, or a change in system behavior may be
the manifestation of a long-term transient. In none of these cases would
CSD be expected to precede such changes. Nonetheless, it may be
difficult to distinguish such cases from bifurcations.

@@

The future of early warning signals lies in the uncharted territory. For
certain classes of transitions, such as stochastically-driven regime
shifts, prediction may not be possible. In such cases, managing for
resilience may be the only option. Likewise, regime shifts driven by
external perturbation or strong forcing are not predictable \emph{if}
the scope of management does not include the external causes. Proper
scoping of the management problem can avoid this situation (Fischer et
al. 2009, Alliance 2010, Polasky et al. 2011). More research is needed
in methods of distinguishing such cases from those in which early
detection may be possible.
